---
output:
  github_document
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# taxinfo <a href="https://adrientaudiere.github.io/taxinfo/"><img src="man/figures/logo.svg" align="right" height="139" alt="taxinfo website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/adrientaudiere/taxinfo/workflows/R-CMD-check/badge.svg)](https://github.com/adrientaudiere/taxinfo/actions)
[![Codecov test coverage](https://codecov.io/gh/adrientaudiere/taxinfo/branch/master/graph/badge.svg)](https://app.codecov.io/gh/adrientaudiere/taxinfo?branch=master)

<!-- [![CRAN status](https://www.r-project.org/Badges/version/taxinfo)](https://CRAN.R-project.org/package=taxinfo)-->
<!--[![CRAN Downloads](https://cranlogs.r-pkg.org/badges/taxinfo)](https://cran.r-project.org/package=taxinfo)-->
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

The **taxinfo** package provides comprehensive tools for augmenting phyloseq objects with taxonomic-based information from various external data sources. It seamlessly integrates data from GBIF, Wikipedia, GLOBI, OpenAlex, TAXREF, and other databases to enrich your taxonomic analyses.

## Overview

**taxinfo** is designed to work with phyloseq objects and provides functions to:

- **Verify and clean taxonomic names** using the Global Names Architecture (GNA)
- **Retrieve occurrence data** from GBIF and other biodiversity databases
- **Access taxonomic traits** from various databases including FungalTraits
- **Get Wikipedia information** including page views, links, and content statistics
- **Fetch scientific literature data** from OpenAlex
- **Access interaction data** from GLOBI (Global Biotic Interactions)
- **Validate geographic occurrences** against ecoregions and biogeographic regions
- **Retrieve taxonomic photos** and media information

## Installation

You can install the stable version of taxinfo from CRAN:

```{r, eval=FALSE}
install.packages("taxinfo")
```

Or the development version from GitHub:

```{r, eval=FALSE}
# Install from GitHub
devtools::install_github("adrientaudiere/taxinfo")

# Or using pak
pak::pkg_install("adrientaudiere/taxinfo")
```

## Key Features

### 🔍 **Data Verification & Quality Control**
- `gna_verifier_pq()`: Verify and standardize taxonomic names using Global Names Architecture

### 🌍 **Biodiversity Data Integration**
- `tax_gbif_occur_pq()`: Retrieve GBIF occurrence data
- `tax_globi_pq()`: Access species interaction data from GLOBI
- `tax_info_pq()`: Add information from CSV files (TAXREF, traits databases)

### 📚 **Knowledge Base Integration**
- `tax_get_wk_info_pq()`: Get comprehensive Wikipedia data
- `tax_oa_pq()`: Retrieve scientific literature from OpenAlex

### 🗺️ **Geographic Analysis**
- `range_bioreg_pq()`: Analyze biogeographic ranges
- `plot_tax_gbif_pq()`: Create distribution maps
- `tax_check_ecoregion()`: Validate occurrences against ecoregions

### 🔬 **Advanced Analysis Tools**
- `tax_retroblast_pq()`: Sequence-based taxonomic verification
- `tax_photos_pq()`: Access taxonomic images and media
- `tax_occur_check_pq()`: Multi-source occurrence validation

### 🎯 **Flexible Input Options**
Most functions can work with either:
- **Phyloseq objects**: Automatically enriches the tax_table (default behavior)
- **Taxonomic name vectors**: Returns tibbles for standalone queries

## Quick Start
### Glossary 

Following the [Darwin core standards](https://dwc.tdwg.org/terms/), here are some key terms used in taxinfo (camel case naming convention):
- **scientificName**: The full scientific name with authorship and date information if known (e.g., "Stereum ostrea (Blume & T.Nees) Fr., 1838")
- **genusEpithet**: Just the genus part (e.g., "Stereum"). Note that the correct Darwin core term is `genus`, but in taxinfo we use `genusEpithet` to avoid confusion with the `genus` field from phyloseq which is often used for the full scientific name.
- **specificEpithet**: Just the species epithet part (e.g., "ostrea")
- **namePublishedInYear**: The year the name was published (e.g., "1838")

Other term come from [verifier globalnames](https://verifier.globalnames.org/) (camel case naming convention):
- **currentCanonicalSimple**: The simplified scientific name without authorship (e.g., "Stereum ostrea"). It correspond to concatenation of the `genusEpithet` and `specificEpithet` fields.

- We add the terms **genusSpeciesEpithet**: Same as `currentCanonicalSimple` but `NA` for genus-only names (i.e. when `specificEpithet` is absent). Useful when you need to filter out unidentified-to-species taxa.

### Example Workflow

```{r}
library(taxinfo)
library(MiscMetabar)

# Load example data (fungal phyloseq object from MiscMetabar)
data("data_fungi_mini", package = "MiscMetabar")

# Step 1: Verify and clean taxonomic names
data_fungi_clean <- gna_verifier_pq(data_fungi_mini,
  data_sources = 210
)

# Step 2: Add GBIF occurrence data (add_to_phyloseq defaults to TRUE)
data_with_gbif <- tax_gbif_occur_pq(data_fungi_clean)

# Step 3: Add trait information
fungal_traits <- system.file("extdata", "fun_trait_mini.csv", package = "taxinfo")
data_with_traits <- tax_info_pq(data_with_gbif,
  taxonomic_rank = "genusEpithet",
  file_name = fungal_traits,
  csv_taxonomic_rank = "GENUS",
  col_prefix = "ft_",
  sep = ";"
)

# Step 4: Add Wikipedia information (add_to_phyloseq defaults to TRUE)
data_final <- tax_get_wk_info_pq(data_with_traits)

# View the enriched taxonomic table
head(data_final@tax_table)

# Alternative: Query specific taxa without a phyloseq object
taxa_info <- tax_gbif_occur_pq(
  taxnames = c("Amanita muscaria", "Boletus edulis"),
  by_country = TRUE
)

# Returns a tibble instead of phyloseq object
head(taxa_info)
```

## Data Sources

**taxinfo** integrates with multiple authoritative data sources:

| Source | Description | Functions |
|--------|-------------|-----------|
| **GBIF** | Global biodiversity occurrence data | `tax_gbif_occur_pq()`, `plot_tax_gbif_pq()` |
| **Wikipedia** | Encyclopedia data and page statistics | `tax_get_wk_info_pq()`, `tax_get_wk_pages_info()` |
| **GLOBI** | Species interaction networks | `tax_globi_pq()` |
| **OpenAlex** | Scientific literature database | `tax_oa_pq()` |
| **TAXREF** | French national taxonomic reference | `tax_info_pq()` |
| **GNA** | Global Names Architecture for name verification | `gna_verifier_pq()` |
| **Custom CSV** | Any taxonomic database in CSV format | `tax_info_pq()` |

## Contributing

We welcome contributions! Please open an issue or pull request on [GitHub](https://github.com/adrientaudiere/taxinfo).

## Citation

If you use taxinfo in your research, please cite:

```r
citation("taxinfo")
```

## Related Packages

**taxinfo** works seamlessly with:

- [MiscMetabar](https://github.com/adrientaudiere/MiscMetabar): Miscellaneous functions for metabarcoding analysis
- [phyloseq](https://joey711.github.io/phyloseq/): Analyze microbiome census data
- [taxize](https://github.com/ropensci/taxize): Taxonomic information from around the web
- [rgbif](https://github.com/ropensci/rgbif): Interface to GBIF API

## Licence

This project is licensed under the MIT License - see the [LICENSE](LICENSE) file for details.


## Acknowledgments

- All the members of the DEFIS MITI project, especially Mélanie Roy and Benoît Perez-Lamarque who lead the project.

- This project has received financial support from the **CNRS** (*Centre National de la Recherche Scientifique*) through the MITI interdisciplinary programs (Project ***DEFIS*** - Exploration of Evolutionary Diversity of Fungi and its Indicators through High-Throughput Sequencing : from multi-actors challenges to long-term monitoring).

- The developers of the R packages used in this project. A special thanks to Joey McMurdie (`phyloseq`), John Waller (`rgbif`), and Zachary Foster (`taxize`) for maintaining those useful tools.
