---
title: "Adding External Information: Wikipedia, GLOBI, and Custom Databases"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Adding External Information: Wikipedia, GLOBI, and Custom Databases}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

```{r setup}
library(taxinfo)
library(MiscMetabar)
library(ggplot2)
library(tidyverse)
```

## Overview

Beyond occurrence data, `taxinfo` integrates multiple knowledge sources to enrich your taxonomic data with Species salience in public knowledge (**Wikipedia**), species interactions (**GLOBI**), scientific literature (**OpenAlex**), and custom database content. This vignette demonstrates how to leverage these functions to create comprehensive taxonomic `@tax_table`.

## Core External Information Functions

- `tax_get_wk_info_pq()`: Wikipedia data and page statistics
- `tax_globi_pq()`: Species interaction data from GLOBI
- `tax_oa_pq()`: Scientific literature from OpenAlex  
- `tax_info_pq()`: Custom database integration (TAXREF, traits, etc.)
- `tax_photos_pq()`: Taxonomic images and media

**Note:** All these functions can work with either phyloseq objects or vectors of taxonomic names (`taxnames` parameter). When using a phyloseq object, results are automatically added to the tax_table. When using `taxnames`, a tibble is returned.

## Wikipedia Integration

Wikipedia is a rich source of public interest and knowledge about species. By integrating Wikipedia data, you can have a proxy of the societal attention given to different taxa. The number of page views, the mean page length and the number of languages in which a species is covered can provide insights into its cultural significance measuring a sort of species salience in public knowledge.



##  Verify and Clean Taxonomic Names

```{r}
# Clean names first
# Keep only first 20 taxa for speed
data_clean <- prune_taxa(taxa = taxa_names(data_fungi_mini)[1:20], data_fungi_mini) |>
  gna_verifier_pq(data_sources = 210)
```


## Wikipedia Integration

### Basic Wikipedia Data

```{r, fig.alt="Lollipop chart showing Wikipedia page views for different taxa on a log scale. Taxa are ordered by page views on the y-axis. Points are sized by page length and colored by Order. Text labels show the number of languages in which each species has a Wikipedia page."}
# Add Wikipedia information (add_to_phyloseq defaults to TRUE for phyloseq objects)
data_clean_wk <- tax_get_wk_info_pq(data_clean, n_days = 30)
# View Wikipedia columns
data_clean_wk@tax_table |>
  as.data.frame() |>
  distinct(currentCanonicalSimple, lang, page_views, Order, page_length) |>
  filter(!is.na(currentCanonicalSimple), !is.na(page_views)) |>
  mutate(across(c(page_views, page_length, lang), as.numeric)) |>
  filter(page_views > 0) |>
  mutate(currentCanonicalSimple = factor(currentCanonicalSimple)) |>
  ggplot(aes(
    y = forcats::fct_reorder(currentCanonicalSimple, page_views),
    x = page_views, size = page_length, col = Order
  )) +
  geom_segment(aes(xend = 0, yend = currentCanonicalSimple), linewidth = 1) +
  geom_point() +
  geom_text(aes(label = lang), size = 3, color = "black") +
  scale_x_log10() +
  labs(
    title = "Species salience in public knowledge",
    x = "Number of page views during a month (log10 scale)",
    y = "Taxa (color=Order)",
    size = "Mean page length"
  ) +
  theme_idest()
```

### Multilingual Wikipedia Analysis

Analyze Wikipedia coverage across languages:

```{r, fig.alt="Scatter plot showing Wikipedia page views (log scale) versus number of Wikipedia language versions for taxa across four languages (en, fr, de, es). Points are sized by mean page length and labeled with taxonomic names."}
data_clean_wk2 <- tax_get_wk_info_pq(data_clean,
  languages_pages = c("en", "fr", "de", "es")
)

# Visualize language coverage
wiki_analysis <- data_clean_wk2@tax_table |>
  as.data.frame() |>
  select(currentCanonicalSimple, lang, page_length, page_views) |>
  mutate(across(c(lang, page_length, page_views), as.numeric)) |>
  filter(!is.na(lang)) |>
  distinct()

ggplot(wiki_analysis, aes(
  x = lang, y = log10(page_views + 1),
  size = page_length
)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  labs(
    title = "Species salience in public knowledge",
    subtitle = "Page views in 4 different countries : en, fr, de and es",
    x = "Number of Wikipedia languages",
    y = "Page views (log10 scale)",
    size = "Mean page length"
  ) +
  theme_idest() +
  ggrepel::geom_text_repel(aes(label = currentCanonicalSimple),
    size = 3,
    max.overlaps = 10,
  )
```


## Species Interactions with GLOBI

### Basic Interaction Data

GLOBI (Global Biotic Interactions) provides data on species interactions including predation, parasitism, pollination, and more (see `rglobi::get_interaction_types()` for all available interaction types). Here we will add interaction data focusing on parasitic and pathogenic relationships.

```{r}
# Get interaction data from GLOBI
data_clean_globi <- tax_globi_pq(data_clean,
  interaction_types = c("parasiteOf", "pathogenOf"),
  max_interactions = 100
)

# View interaction columns
head(data_clean_globi@tax_table[, c(
  "nb",
  "target_taxon_name",
  "pathogenOf",
  "parasiteOf"
)])
```

```{r, fig.alt="Scatter plot showing the number of interactions in GLOBI for different fungal taxa. Points are colored by ecological guild and sized by molecular abundance (log scale). Text labels show the count of pathogen and parasite host relationships."}
psmelt(data_clean_globi) |>
  group_by(taxa_name) |>
  summarise(
    Abundance = sum(Abundance),
    nb_inter = mean(map_dbl(nb, ~ sum(as.numeric(unlist(strsplit(.x, "; "))), na.rm = TRUE)), na.rm = TRUE),
    n_host_pathog = mean(map_dbl(pathogenOf, ~ stringr::str_count(.x, ";")), na.rm = TRUE),
    n_host_parasit = mean(map_dbl(parasiteOf, ~ stringr::str_count(.x, ";")), na.rm = TRUE),
    Guild = Guild[1]
  ) |>
  mutate(n_host_parasit = ifelse(is.na(n_host_parasit), 0, n_host_parasit)) |>
  mutate(n_host_pathog = ifelse(is.na(n_host_pathog), 0, n_host_pathog)) |>
  filter(nb_inter > 0) |>
  ggplot(aes(
    y = forcats::fct_reorder(taxa_name, nb_inter),
    x = nb_inter,
    color = Guild,
    size = log10(1 + Abundance)
  )) +
  geom_point() +
  geom_text(aes(label = paste(n_host_pathog, "-", n_host_parasit)),
    size = 2.5, color = "black", nudge_y = 0.2
  ) +
  theme_idest() +
  labs(
    title = "Interactions in GLOBI",
    subtitle = "First and second number indicate the number of verified taxonomic\nentity whose the taxa are respectively pathogen or parasite.",
    x = "Number of interactions",
    y = "Taxa",
    color = "Guild following FunGuild",
    size = "Number of sequences (log10)"
  )
```

### Detailed Interaction Analysis

Get detailed interaction data for further analysis:

```{r, fig.alt="Violin plot with jittered points showing the distribution of number of target taxa by interaction type (hasHost, parasiteOf, other). Each violin is filled by interaction type."}
# Get detailed interaction tibble (not added to phyloseq)
detailed_interactions <- tax_globi_pq(data_clean,
  interaction_types = c("parasiteOf", "hasHost", "pathogenOf"),
  max_interactions = 100,
  add_to_phyloseq = FALSE
)

# Analyze interaction patterns
interaction_summary <- detailed_interactions |>
  separate_rows(target_taxon_name, nb, sep = ";\\s*", convert = TRUE) |>
  mutate(across(
    all_of(c("hasHost", "parasiteOf")),
    ~ stringr::str_detect(.x, stringr::fixed(target_taxon_name))
  )) |>
  mutate(interaction_type = case_when(
        parasiteOf ~ "parasiteOf",
    hasHost ~ "hasHost",
    .default = "other"
  ))

# Visualize interaction networks
ggplot(interaction_summary, aes(
  x = interaction_type, y = nb,
  fill = interaction_type
)) +
  geom_violin() +
  geom_jitter(col = "grey40", alpha = 0.2, height = 0) +
  labs(
    title = "Species Interaction Diversity",
    x = "Interaction Type",
    y = "Number of Target Taxa"
  ) +
  theme_idest() +
  theme(legend.position = "none")
```

```{r, fig.alt="Network diagram showing fungal-plant interactions as segments connecting fungal taxa (darkred points at bottom) to plant target taxa (darkgreen points at top). Segments are colored by interaction type and sized by number of interactions."}
ggplot(
  interaction_summary,
  aes(x = taxa_name, y = target_taxon_name)
) +
  geom_segment(
    aes(
      x = taxa_name, xend = target_taxon_name,
      y = 0, yend = 1,
      color = interaction_type,
      linewidth = nb
    ),
    alpha = 0.6
  ) +
  geom_point(aes(x = taxa_name, y = 0), size = 3, color = "darkred") +
  geom_point(aes(x = target_taxon_name, y = 1), size = 3, color = "darkgreen") +
  scale_linewidth_continuous(range = c(0.5, 3), name = "Number of interactions") +
  scale_color_viridis_d(name = "Interaction type") +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom"
  ) +
  coord_flip() +
  labs(title = "Fungal-Plant Interaction Network")
```

```{r, fig.alt="Network graph using Fruchterman-Reingold layout showing fungal-plant interaction network. Fungal nodes are shown in orange and plant nodes in light green. Edges are colored by interaction type and sized by number of interactions. Only fungal taxa names are labeled."}
library(ggraph)

edges <- interaction_summary |>
  select(
    from = taxa_name, to = target_taxon_name,
    weight = nb, type = interaction_type
  )

nodes <- data.frame(
  name = unique(c(edges$from, edges$to))
) |>
  mutate(
    node_type = ifelse(name %in% unique(edges$from), "Fungal", "Plant")
  )

graph <- igraph::graph_from_data_frame(edges, directed = FALSE, vertices = nodes)

ggraph(graph, layout = "fr") +
  geom_edge_link(aes(width = weight, color = type), alpha = 0.6) +
  geom_node_point(aes(fill = node_type), size = 5, shape = 21, color = "black") +
  geom_node_text(aes(label = ifelse(node_type == "Fungal", name, "")), repel = TRUE, size = 3) +
  scale_edge_width_continuous(range = c(0.5, 3), name = "Number of interactions") +
  scale_edge_color_manual(values = c("grey60", "purple", "cyan", "green")) +
  scale_fill_manual(values = c("Plant" = "lightgreen", "Fungal" = "orange")) +
  theme_idest(grid = FALSE, axis_text_size = 0) +
  labs(
    title = "Fungal-Plant Interaction Network",
    y = "",
    x = ""
  )
```


## Scientific Literature with OpenAlex

### Literature Metrics

Get publication data for your taxa and add it to the previous wikipedia-enhanced dataset:

```{r}
data_clean_oa <- tax_oa_pq(data_clean_wk)

head(data_clean_oa@tax_table[, c(
  "n_doi",
  "list_doi",
  "taxa_name"
)])
```

### Research Interest Analysis

Analyze research patterns:

```{r, fig.alt="Scatter plot comparing scientific interest (number of publications, log scale) to public interest (Wikipedia page views, log scale). Points are sized by mean citations per publication and colored by number of Wikipedia languages. A linear regression line with equation and R-squared value is shown. Taxonomic names are labeled."}
data_clean_oa@tax_table |>
  as.data.frame() |>
  select(
    currentCanonicalSimple,
    n_doi,
    n_citation,
    page_views,
    lang,
    Family
  ) |>
  mutate(across(any_of(c("n_doi", "n_citation", "page_views", "lang")), as.numeric)) |>
  distinct(currentCanonicalSimple, .keep_all = TRUE) |>
  ggplot(aes(
    x = log10(n_doi + 1),
    y = log10(page_views + 1)
  )) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_point(aes(size = n_citation / n_doi, color = lang), alpha = 0.6) +
  labs(
    title = "Scientific Interest vs Public Interest",
    x = "Number of Publications (log10)",
    y = "Wikipedia Page Views (log10)",
    size = "Mean nb of citation"
  ) +
  ggrepel::geom_text_repel(aes(label = currentCanonicalSimple), size = 3, color = "black", fontface = "italic") +
  theme_idest() +
  ggpmisc::stat_poly_eq(
    aes(label = paste(..eq.label.., ..rr.label.., sep = "~~~")),
    formula = y ~ x,
    parse = TRUE
  )
```

## Custom Database Integration

### Fungal Traits Database

Integrate trait databases using `tax_info_pq()`. Here we will use a bigger 
phyloseq object to found more informations in the mini-trait database. When 
using true database, you should find more information, even with less taxa.

```{r}
data_clean_full <- gna_verifier_pq(data_fungi, data_sources = 210)

# Load fungal traits database
fungal_traits <- system.file("extdata", "fun_trait_mini.csv",
  package = "taxinfo"
)

# Add trait information
data_clean_ft <- tax_info_pq(data_clean_full,
  taxonomic_rank = "genus",
  file_name = fungal_traits,
  csv_taxonomic_rank = "GENUS",
  col_prefix = "ft_",
  sep = ";"
)

# View trait columns
data_clean_ft@tax_table |>
  as.data.frame() |>
  pull(ft_primary_lifestyle) |>
  table()
```


### Example Trait Visualization

Here's an example of fungal livestyle distribution

```{r echo=FALSE, fig.alt="Horizontal bar chart showing fungal lifestyle distribution. Bars are colored by Guild (e.g., saprotroph, plant pathogen) and ordered by occurrence (number of samples). Text labels show molecular abundance for each taxon.", out.width="100%"}
psmelt(data_clean_ft) |>
  group_by(currentCanonicalSimple) |>
  summarise(
    lifestyle = unique(ft_primary_lifestyle),
    occurence = sum(Abundance > 0, na.rm = TRUE),
    Abundance = sum(Abundance, na.rm = T),
    Family = unique(Family)
  ) |>
  filter(!is.na(lifestyle)) |>
  ggplot(aes(
    x = forcats::fct_reorder(currentCanonicalSimple, occurence),
    y = 1 + occurence,
    fill = lifestyle
  )) +
  geom_col() +
  geom_text(aes(label = Abundance, y = 2 + occurence), size = 3, color = "black", hjust = 0) +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Fungal Lifestyle",
    subtitle = "Number in black indicates the total number of sequences for each taxa.",
    x = "Primary Lifestyle",
    y = "Number of occurrence (samples)"
  ) +
  theme_idest() +
  theme(axis.text.y = element_text(face = "italic"))
```

### TAXREF Integration

For French taxonomic data, integrate TAXREF:

```{r}
# Load TAXREF data (example file)
taxref_file <- system.file("extdata", "TAXREFv18_fungi.csv",
  package = "taxinfo"
)

# Add TAXREF information
data_clean_taxref <- tax_info_pq(data_clean,
  file_name = taxref_file,
  csv_taxonomic_rank = "LB_NOM",
  csv_cols_select = NULL,
  col_prefix = "taxref_"
)
```

```{r, fig.alt="Horizontal bar chart showing number of sample of taxa. Bars are colored by Order. Taxa known in France (i.e. present in TAXREF) are shown in bold. French vernacular names are shown as text labels."}
psm <- psmelt(data_clean_taxref) |>
  group_by(currentCanonicalSimple) |>
  mutate(across(everything(), ~ replace(., . == "" | . == "NA", NA))) |>
  filter(!is.na(currentCanonicalSimple)) |>
  summarise(
    taxref_FR = unique(taxref_FR),
    taxref_HABITAT = unique(taxref_HABITAT),
    occurence = sum(Abundance > 0, na.rm = TRUE),
    Abundance = sum(Abundance, na.rm = T),
    Order = unique(taxref_ORDRE),
    taxref_NOM_VERN = unique(taxref_NOM_VERN)
  ) |>
  filter(!is.na(Order))

ggplot(psm, aes(
  x = forcats::fct_reorder(currentCanonicalSimple, occurence),
  y = 1 + occurence,
  fill = Order
)) +
  geom_col() +
  geom_text(aes(label = currentCanonicalSimple,
                fontface = ifelse(is.na(taxref_FR), "italic", "bold.italic")),
    y = -1, hjust = 1, size = 2.5
  ) +
  geom_text(aes(label = taxref_NOM_VERN, y = 2 + occurence), size = 3, color = "black", hjust = 0) +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "Frequences of Taxa with French common names",
    subtitle = "Bold taxa are already known in France (TAXREF).",
    x = "Taxa",
    y = "Number of occurences (samples)"
  ) +
  theme_idest() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.35, 0.05)))
```


## Comprehensive Data Integration

### Multi-Source Enrichment

Combine multiple data sources for comprehensive profiles:

```{r}
# Complete multi-source enrichment
data_fungi_full <- data_fungi_mini |>
  gna_verifier_pq(data_sources = 210) |>
  tax_gbif_occur_pq() |>
  tax_get_wk_info_pq() |>
  tax_oa_pq(count_only = TRUE) |>
  tax_globi_pq(
    interaction_types = c("parasiteOf", "pathogenOf"),
    max_interactions = 20
  ) |>
  tax_info_pq(
    file_name = taxref_file,
    csv_taxonomic_rank = "LB_NOM",
    csv_cols_select = NULL,
    col_prefix = "taxref_"
  )

# View enriched tax_table
data_fungi_full@tax_table[1:3, ]
```

### Data Quality Assessment

Assess information completeness across sources:

```{r}
# Calculate information completeness
completeness_analysis <- data_fungi_full@tax_table |>
  as.data.frame() |>
  summarise(
    gbif = 100 * mean(!is.na(Global_occurences)),
    wikipedia = 100 * mean(!is.na(lang)),
    globi = 100 * mean(!is.na(nb)),
    taxref = 100 * mean(!is.na(taxref_CD_NOM)),
    openalex = 100 * mean(!is.na(n_doi))
  ) |>
  tidyr::pivot_longer(everything(), names_to = "data_source", values_to = "completeness")
```

```{r, fig.alt="Horizontal bar chart showing data completeness percentages for different external data sources (GBIF, Wikipedia, GLOBI, TAXREF, OpenAlex). Each bar shows the percentage of taxa with available data from that source, with percentage labels in the middle of each bar."}
# Visualize data completeness
ggplot(completeness_analysis, aes(
  x = reorder(data_source, completeness),
  y = completeness, fill = data_source
)) +
  geom_col() +
  geom_hline(yintercept = 100) +
  coord_flip() +
  geom_label(aes(label = paste0(round(completeness, 1), "%"), y = completeness / 2), hjust = 0.5, col = "black", fill = rgb(1, 1, 1, 0.5)) +
  scale_fill_viridis_d() +
  labs(
    title = "Data Source Completeness",
    x = "Data Source",
    y = "Percentage of Taxa with Data"
  ) +
  theme_idest() +
  ylim(c(0, 100)) +
  theme(legend.position = "none")
```

### Integration Visualization

Create comprehensive visualization of integrated data:

```{r, fig.alt="Scatter plot showing GBIF occurrences (x-axis, log scale) versus Wikipedia page views (y-axis, log scale). Points are colored by Order, shaped by presence in France (TAXREF), and sized by number of publications (log scale)."}
# Prepare data for visualization
viz_data <- data_fungi_full@tax_table |>
  as.data.frame() |>
  mutate(taxref_FR = tidyr::replace_na(taxref_FR, "")) |>
  mutate(
    Global_occurences = as.numeric(Global_occurences),
    wk_sum_page_views = as.numeric(page_views),
    globi_nb_interactions = as.numeric(nb),
    oa_n_doi = as.numeric(n_doi),
    taxref = taxref_FR != ""
  ) |>
  filter(!is.na(Global_occurences) | is.na(wk_sum_page_views)) |>
  distinct(currentCanonicalSimple, .keep_all = TRUE)

# Multi-dimensional visualization
ggplot(viz_data, aes(
  x = log10(Global_occurences + 1),
  y = log10(wk_sum_page_views + 1)
)) +
  geom_point(
    aes(
      size = log10(1 + as.numeric(oa_n_doi)),
      shape = taxref,
      color = Order
    ),
    alpha = 0.8
  ) +
  scale_shape_manual(values = c(17, 16), name = "Presence in France") +
  scale_color_viridis_d(name = "Order") +
  labs(
    title = "Integrated Taxonomic Information",
    subtitle = "GBIF occurrences, Wikipedia popularity, interactions, and traits",
    x = "GBIF Occurrences (log10)",
    y = "Wikipedia Page Views (log10)",
    size = "n_doi (log10)"
  ) +
  ggrepel::geom_text_repel(aes(label = currentCanonicalSimple), size = 3, fontface = "italic") +
  theme_idest()
```




### Knowledge Gap Identification

Identify taxa with limited information:

```{r}
# Identify knowledge gaps
knowledge_gaps <- data_fungi_full@tax_table |>
  as.data.frame() |>
  select(Global_occurences, lang, nb, taxref_CD_NOM, n_doi, currentCanonicalSimple) |>
  mutate(
    gbif_info = !is.na(Global_occurences),
    wk_info = !is.na(lang),
    oa_info = !is.na(n_doi),
    taxref_info = (!is.na(taxref_CD_NOM) | taxref_CD_NOM == "NA"),
    globi_info = (!is.na(nb))
  ) |>
  mutate(taxref_info = ifelse(is.na(taxref_info), FALSE, taxref_info)) |>
  mutate(
    info_score = gbif_info + wk_info + oa_info + taxref_info + globi_info
  ) |>
  arrange(info_score)

# Number of available sources of information per taxa (ASV here)
knowledge_gaps |>
  pull(info_score) |>
  table()


# Number of available sources of information per taxonomic names (current canonical simple)
knowledge_gaps |>
  distinct(currentCanonicalSimple, .keep_all = TRUE) |>
  pull(info_score) |>
  table()

# Poorly knowns taxonomic names (current canonical simple)
knowledge_gaps |>
  distinct(currentCanonicalSimple, .keep_all = TRUE) |>
  filter(info_score <= 2) |>
  pull(currentCanonicalSimple)
```

```{r, fig.alt="UpSet plot showing intersections of data availability across five information sources (GBIF, Wikipedia, OpenAlex, TAXREF, GLOBI). Horizontal bars show the size of each data source, and vertical bars show the size of unique combinations of sources for different taxa."}
knowledge_gaps |>
  distinct(currentCanonicalSimple, .keep_all = TRUE) |>
  filter(!is.na(currentCanonicalSimple)) |>
  ComplexUpset::upset(
    intersect = c("gbif_info", "wk_info", "oa_info", "taxref_info", "globi_info"),
    keep_empty_groups = TRUE,
    wrap=TRUE,set_sizes = F
      
  ) + 
  labs(
    title = "Knowledge Gaps Across Data Sources",
    subtitle = "Number of taxa names with available data from each source and their intersections"
  ) + theme_idest()
```


## Best Practices

When integrating external data, consider the following best practices:
1. **Taxonomic matching**: Ensure consistent naming across sources
2. **Data validation**: Check for outliers or errors
3. **Missing data**: Handle NAs appropriately in analyses
4. **Documentation**: Keep track of data sources and versions

This comprehensive approach to external data integration transforms basic taxonomic lists into rich, multi-dimensional datasets suitable for advanced ecological analyses.

