---
title: "Getting Started with taxinfo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with taxinfo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figures/",
  out.width = "100%"
)
```

```{r setup}
library(taxinfo)
library(MiscMetabar)
library(ggplot2)
library(dplyr)
```

## Overview

The **taxinfo** package provides comprehensive tools for augmenting phyloseq objects with taxonomic-based information from various external data sources. This vignette provides a big picture overview of the package's capabilities and workflow.

## Key Features

**taxinfo** integrates with multiple authoritative data sources to enrich your taxonomic data:

- **🔍 Data Verification**: Verify and standardize taxonomic names using Global Names Architecture
- **🌍 Biodiversity Data**: Access GBIF occurrence data and species interactions from GLOBI  
- **📚 Knowledge Integration**: Retrieve Wikipedia data and scientific literature from OpenAlex
- **🗺️ Geographic Analysis**: Analyze biogeographic ranges and create distribution maps
- **🔬 Advanced Tools**: Sequence-based verification and multi-source occurrence validation

## Main Data Sources

| Source | Description | Key Functions |
|--------|-------------|---------------|
| **GBIF** | Global biodiversity occurrence data | `tax_gbif_occur_pq()`, `plot_tax_gbif_pq()` |
| **Wikipedia** | Encyclopedia data and page statistics | `tax_get_wk_info_pq()` |
| **GLOBI** | Species interaction networks | `tax_globi_pq()` |
| **OpenAlex** | Scientific literature database | `tax_oa_pq()` |
| **TAXREF** | French national taxonomic reference | `tax_info_pq()` |
| **GNA** | Global Names Architecture for name verification | `gna_verifier_pq()` |
| **Custom CSV** | Any taxonomic database in CSV format | `tax_info_pq()` |

## Basic Workflow

### Step 1: Load Example Data

```{r}
# Load example fungal data from MiscMetabar
data("data_fungi_mini", package = "MiscMetabar")

# Check the structure
data_fungi_mini
```

### Step 2: Verify and Clean Taxonomic Names

The first step is typically to verify and standardize taxonomic names using the Global Names Architecture:

```{r eval=FALSE}
# Verify and clean taxonomic names
data_clean <- gna_verifier_pq(data_fungi_mini, 
                             data_sources = 210,  # TaxRef
                             add_to_phyloseq = TRUE)

# View the enhanced taxonomic table
head(data_clean@tax_table)
```

This adds standardized columns:
- `submittedName`: Original name submitted
- `currentName`: Current accepted name with authorities  
- `currentCanonicalSimple`: Clean accepted name without authorities

### Step 3: Add Biodiversity Information

Once names are verified, you can enrich your data with various sources:

```{r eval=FALSE}
# Add GBIF occurrence data
data_with_gbif <- tax_gbif_occur_pq(data_clean,
                                   add_to_phyloseq = TRUE)

# Add species interaction data from GLOBI
data_with_globi <- tax_globi_pq(data_with_gbif,
                               interaction_types = "hasHost",
                               add_to_phyloseq = TRUE)

# Add Wikipedia information
data_enriched <- tax_get_wk_info_pq(data_with_globi,
                                   add_to_phyloseq = TRUE)
```

### Step 4: Add Custom Database Information

You can also integrate custom databases or trait information:

```{r eval=FALSE}
# Add fungal trait information
fungal_traits <- system.file("extdata", "fun_trait_mini.csv", 
                            package = "taxinfo")

data_final <- tax_info_pq(data_enriched,
                         taxonomic_rank = "currentCanonicalSimpleGenus",
                         file_name = fungal_traits,
                         csv_taxonomic_rank = "GENUS",
                         col_prefix = "ft_",
                         sep = ";")
```

## Function Categories

### Data Verification & Quality Control
- `gna_verifier_pq()`: Verify and standardize taxonomic names

### Biodiversity Data Integration  
- `tax_gbif_occur_pq()`: Retrieve GBIF occurrence data
- `tax_globi_pq()`: Access species interaction data from GLOBI
- `tax_info_pq()`: Add information from CSV files

### Knowledge Base Integration
- `tax_get_wk_info_pq()`: Get comprehensive Wikipedia data
- `tax_oa_pq()`: Retrieve scientific literature from OpenAlex

### Geographic Analysis
- `range_bioreg_pq()`: Analyze biogeographic ranges
- `plot_tax_gbif_pq()`: Create distribution maps
- `tax_check_ecoregion()`: Validate occurrences against ecoregions

### Advanced Analysis Tools
- `tax_retroblast_pq()`: Sequence-based taxonomic verification
- `tax_photos_pq()`: Access taxonomic images and media
- `tax_occur_check_pq()`: Multi-source occurrence validation

## Best Practices

1. **Always start with name verification** using `gna_verifier_pq()`
2. **Use appropriate data sources** for your taxonomic group
3. **Set reasonable limits** for API calls to avoid timeouts
4. **Cache results** for large datasets to avoid repeated API calls
5. **Validate geographic occurrences** using occurrence checking functions

## Next Steps

- **Checking Taxa Presence**: Learn how to validate if taxa are likely present in your samples
- **GBIF-based Functions**: Explore occurrence data and distribution mapping  
- **Adding External Information**: Integrate Wikipedia, GLOBI, and custom databases

This provides the foundation for using taxinfo effectively. Each subsequent vignette will dive deeper into specific functionality areas.