---
title: "Getting Started with taxinfo: Augment phyloseq Objects with Taxonomic Information"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    code_folding: show
    theme: united
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Getting Started with taxinfo}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

## Overview

The `taxinfo` package is designed to augment phyloseq objects with additional taxonomic-based information. This package provides tools to enhance your microbiome analysis by enriching taxonomic data with external database information, functional annotations, and taxonomic lineage details.

## Key Features

- Augment phyloseq objects with comprehensive taxonomic information
- Retrieve taxonomic data from multiple databases
- Add functional annotations to taxa
- Generate taxonomic lineage information
- Integrate seamlessly with the phyloseq workflow

## Installation

### Installing from GitHub

You can install the development version of `taxinfo` directly from GitHub:

```{r installation, eval=FALSE}
# Install remotes if you haven't already
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install taxinfo from GitHub
remotes::install_github("adrientaudiere/taxinfo")
```

```{r}
library(taxinfo)
```

# Quick Start

## Basic Example
